A Review of Neutron Imaging

### 2020-09-29 neutron-imaging/neutron.tex
WIP29Sep2020

\section{Interaction of neutrons with matter}\label{s4}
Neutrons being electrically neutral can travel significant distances in
condensed matter. The probability $\mu$ that a neutron interacts with matter
in unit distance is called its linear attenuation coefficient. It depends on
the energy of the neutron and the properties of the condensed matter through
which it is traveling. If $N(x)$ is the number of neutron that have traversed
a distance $x$ in the condensed matter without any interaction then the number
of neutrons $dN$ that will interact in the distance interval $x$ and $x + dx$
is
\begin{equation}\label{s4e1}
dN = -\mu Ndx.
\end{equation}
The solution of this equation is 
\begin{equation}\label{s4e2}
N = N_0 e^{-\mu x},
\end{equation}
where $N_0 = N(x = 0)$ is the number of neutrons incident on a beam of the
material. Neutrons interact only with atomic nuclei. In an elastic collision
they transfer a portion of their kinetic energy to the nuclei. A series of
elastic collisions leads to slowing of neutrons, also called their moderation.
Eventually, the kinetic energy of the neutron drops down to the level of the
thermal energy of the surroundings and it gets absorbed by a nucleus. Neutrons
can also undergo inelastic collisions with nuclei. In this case, the neutron
transfers a portion of its energy to the incident nucleus, raising it to an 
excited state. Even in this case, neutron meets the same fate of losing energy
up to the level of thermal energy and getting absorbed. Materials rich in 
protons, that is H${}^+$ ions, show an overwhelmingly large proportion of
elastic collisions because the bare proton does not have excited states. The
de-energized neutron is absorbed leading the formation of a Deuterium atom
\begin{equation}\label{s4e3
{}_1^1H + {}^1_0 n \rightarrow {}^2_1 H + \gamma.
\end{equation}
The energy of the resulting $\gamma$-radiation is $2.22$ MeV. The precise 
energy of this radiation is used to detect the presence of neutrons.
In
materials with other nuclei, a plot of the neutron collision cross sections 
with neutron energies show pronounced peaks and troughs at the nuclear 
transition levels. A notable aspect of elastic collisions is that a neutron of
mass $m_n$ can transfer its kinetic energy to a nucleus of mass $m$ to the
extent of
\begin{equation}\label{s4e4}
Q = \frac{4m_nmE_n}{(m + m_n)^2},
\end{equation}
where $E_n$ is the energy of the neutrons. In a material rich with protons,
$m = m_n$ and $Q = 1$. Light nuclei are thus far more effective in slowing
neutrons than heavier ones. Elastic collisions of neutrons with heavy nuclei
change the direction of motion of nuclei while absorbing only a small portion
of their kinetic energy.
\section{Detection of neutrons}\label{s5}
\section{Neutron dosimetry and shielding}\label{s6}
\section{Propagation of neutron beams}\label{s7}
\section{Neutron diffraction by crystals}\label{s8}
\section{2-d neutron imaging}\label{s9}
\section{Neutron tomography}\label{s10}
\section{Imaging of biologically important structures}\label{s11}
\section{Larmor precession based imaging}\label{s12}
\section{Polarization based imaging}\label{s13}
\section{Conclusion}\label{s14}
\bibliography{neutron}
\bibliographystyle{plain}
\end{document}

>>>>

\section{Interaction of neutrons with matter}\label{s4}
Neutrons being electrically neutral can travel significant distances in
condensed matter. The probability $\mu$ that a neutron interacts with matter
in unit distance is called its linear attenuation coefficient. It depends on
the energy of the neutron and the properties of the condensed matter through
which it is traveling. If $N(x)$ is the number of neutron that have traversed
a distance $x$ in the condensed matter without any interaction then the number
of neutrons $dN$ that will interact in the distance interval $x$ and $x + dx$
is
\begin{equation}\label{s4e1}
dN = -\mu Ndx.
\end{equation}
The solution of this equation is 
\begin{equation}\label{s4e2}
N = N_0 e^{-\mu x},
\end{equation}
where $N_0 = N(x = 0)$ is the number of neutrons incident on a beam of the
material. Neutrons interact only with atomic nuclei. In an elastic collision
they transfer a portion of their kinetic energy to the nuclei. A series of
elastic collisions leads to slowing of neutrons, also called their moderation.
Eventually, the kinetic energy of the neutron drops down to the level of the
thermal energy of the surroundings and it gets absorbed by a nucleus. Neutrons
can also undergo inelastic collisions with nuclei. In this case, the neutron
transfers a portion of its energy to the incident nucleus, raising it to an 
excited state. Even in this case, neutron meets the same fate of losing energy
up to the level of thermal energy and getting absorbed. Materials rich in 
protons, that is H${}^+$ ions, show an overwhelmingly large proportion of
elastic collisions because the bare proton does not have excited states. The
de-energized neutron is absorbed leading the formation of a Deuterium atom
\begin{equation}\label{s4e3}
{}_1^1H + {}^1_0 n \rightarrow {}^2_1 H + \gamma.
\end{equation}
The energy of the resulting $\gamma$-radiation is $2.22$ MeV. The precise 
energy of this radiation is used to detect the presence of neutrons.
In
materials with other nuclei, a plot of the neutron collision cross sections 
with neutron energies show pronounced peaks and troughs at the nuclear 
transition levels. A notable aspect of elastic collisions is that a neutron of
mass $m_n$ can transfer its kinetic energy to a nucleus of mass $m$ to the
extent of
\begin{equation}\label{s4e4}
Q = \frac{4m_nmE_n}{(m + m_n)^2},
\end{equation}
where $E_n$ is the energy of the neutrons. In a material rich with protons,
$m = m_n$ and $Q = 1$. Light nuclei are thus far more effective in slowing
neutrons than heavier ones. Elastic collisions of neutrons with heavy nuclei
change the direction of motion of nuclei while absorbing only a small portion
of their kinetic energy.
\section{Detection of neutrons}\label{s5}
Neutrons are chargeless and their magnetic moment is too low to be used for
reliably detecting their presence. They also have a small electric dipole 
moment but it has not been measured in a laboratory. Free neutrons decay to
produce a proton and an electron but its half life of around $10$ minutes is
too long to permit the detection of neutrons through its decay products. That
leaves us with just one option, consider nuclear reactions triggered by the
neutrons and measure their results to estimate the number and energy of the
neutrons. Since the imaging applications involve mostly slow neutrons, we will
focus our attention on their detection.

Slow neutrons are detected by the nuclear reactions they trigger in ${}^{10}_5$
B, ${}^6_3$Li and ${}^3_2$He. The first two of these produce alpha particles
while the last one produces hydrogen atom. One can use non-relativistic energy
and momentum conservation laws to show that the energy of the alpha particle
in these reactions is \cite{turner2008atoms}
\begin{equation}\label{s5e1}
E = \frac{MQ}{m + M},
\end{equation}
where $m$ is the mass of the alpha particle, $M$ that of the daughter 
nucleus and $Q$ is the energy released in the reaction. The resulting alpha
particles are detected using proportional counters. Proportional counters
measure only the number of neutrons, not their energy spectrum. The reactions
used for detection are \cite{crane1991neutron}
\begin{equation}\label{s5e2}
{}^{10}_5\text{B}+{}^1_0\text{n} \rightarrow \begin{cases}
{}^7_3\text{Li}^\star + {}^4_2\text{He} + 2.31\text{ MeV} \\
{}^7_3\text{Li} + {}^4_2\text{He} + 2.79\text{ MeV}, 
\end{cases}
\end{equation}
where the $\star$ superscript in the first reaction indicates that the nucleus
is in an excited state. The other reactions are
\begin{eqnarray}
{}^6_3\text{Li}+{}^1_0\text{n}&\rightarrow&{}^3_1\text{H} + {}^4_2\text{He} +
4.78\text{ MeV} \label{s5e3} \\
{}^3_2\text{He}+{}^1_0\text{n}&\rightarrow&{}^3_1\text{H} + {}^1_1\text{H} +
0.765\text{ MeV} \label{s5e4} 
\end{eqnarray}
\section{Neutron dosimetry and shielding}\label{s6}
\section{Propagation of neutron beams}\label{s7}
\section{Neutron diffraction by crystals}\label{s8}
\section{2-d neutron imaging}\label{s9}
\section{Neutron tomography}\label{s10}
\section{Imaging of biologically important structures}\label{s11}
\section{Larmor precession based imaging}\label{s12}
\section{Polarization based imaging}\label{s13}
\section{Conclusion}\label{s14}
\bibliography{neutron}
\bibliographystyle{plain}
\end{document}
